Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2020-11-29 GWLDCI.tex
specified citations in paper 1

We now define the notion of an admissible diagram, which imposes certain density and non-crossing conditions on the configuration of the propagators. The motivation for these conditions comes from the link to SYM $N=4$ theory, where the admissible Wilson loop diagrams are exactly those that correspond to $N^kMHV$ diagrams; see for example \cite{Amplituhedronsquared}.

@@

There are a few things to note here.  First, what we call weakly admissible here is called admissible in \cite{wilsonloop}.  \hlfix{Many of the results below still apply when the hypothesis of admissibility is changed to weak admissibility.}{Propagate "weakly" through document}

@@

As described in the introduction, Wilson loop diagrams come from the study of SYM $N = 4$ in the physics literature. In particular (see \cite{Adamo:2012xe} for details) this gives that each diagram $W$ is associated to a matrix of data $C(W)$, which we now describe.

@@

Matroid theory relates to the study of Wilson loop diagrams as follows. In \cite{wilsonloop}, Agarwala and Marin-Amat show that every weakly admissible Wilson loop diagram $W$ with $k$ propagators defines a positroid $M(W)$ of rank $k$ on a base set of size $n$. Combined with Postnikov's work in \cite{Postnikov}, this shows that each admissible Wilson loop diagram corresponds to a convex subset of $\Gr(k,n)$, thus connecting the geometry of Wilson loop diagrams with the geometry underlying the Amplituhedron.

@@

\emph{Polygon dissection} is a common term in combinatorics for a decomposition of a convex polygon into smaller polygons by means of noncrossing diagonals; see for example \cite{PSpolygon}.  Each $\tau(W)$ is such an object.

@@

\begin{proof}
Let us first record a few elementary facts about polygon triangulations (that is, about triangulations with all vertices on the outer face).  If such a triangulation has $n$ vertices then it has $n$ edges on the polygon (that is, on the outer face) and $n-3$ edges which are not.  No planar simple graph with the same vertices and the same outer face can have more edges than a triangulation, and every such simple graph with $n-3$ edges off the outer face is a triangulation.\footnote{The last sentence is the statement that maximal outerplanar graphs are exactly polygon triangulations\cite{MPhD}, then the earlier facts follow from Euler's formula for planar graphs.}

@@

In \cite{wilsonloop}, the authors show that each weakly admissible Wilson loop diagram corresponds to a positroid. That is, each diagram corresponds to a matroid that can be represented by elements of the positive Grassmannian $\Gr(|\cP|, n)$. Any positroid of rank $k$ on $n$ elements defines a subspace of the positive Grassmannian $\Gr(k, n)$, namely the points which represent it \cite{Postnikov}. These subspaces give a CW structure on $\Gr(k,n)$, with each positroid defining a cell.  In other words, there is a map from weakly admissible Wilson loop diagrams $W$ to matroids $M(W)$ to positroid cells in the CW complex on $\Gr(|\cP|, n)$.

@@

Now we are ready to prove the main result of this section, namely that two Wilson loop diagrams define the same matroid (positroid) if and only if they are equivalent (Theorem \ref{same matroid iff equiv}). We also give a formula for the number of Wilson loop diagrams in a given equivalence class (Corollary \ref{number of equiv diagrams}), completing the characterization of the correspondence between Wilson loop diagrams and positroids starting in \cite{wilsonloop}.

@@

We know from examples (e.g. \cite{casestudy}) that many positroid cells cannot be obtained as the matroids of admissible Wilson loop diagrams. We do not obtain a formula for the number of cells in the image of $M$ here, but in the next section we give an alternate characterization of ${\rm im}(M)$ in terms of associahedra, with the hope that this linkage with a well-studied combinatorial object will shed light on the issue in the future.

@@

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  There are many interesting realizations of the associahedron \cite{CSZinequivalent}. The following realization will be the one that we use.

@@

    Since $\tau(W)$ and $\tau(X)$ do not have the same set of non-triangle faces, there must be some diagonal $d$ of $\tau(W)$ that crosses through a non-triangle face $f$ of $\tau(X)$. 
    It will be most convenient to consider faces in terms of their normal vectors. We make use of the construction discussed on page 8 of \cite{CSZinequivalent}: namely, pick one side of $d$ to be positive and define the $i$th coordinate of the vector $\omega^+_d$ to be
    \[
    \omega^+_d(i) = \begin{cases} \text{dist}(x_i, d) & \text{if $x$ is on the positive side of $d$,}\\ 0 & \text{otherwise.} \end{cases}
    \]
    We have from \cite{CSZinequivalent} that $\omega^+_d$ is a normal vector for any face with $d$ as a diagonal.  However we can also check this fact in a completely elementary way using the shoelace formula for the area of a triangle.  The shoelace formula says that if $(x_1, y_1), (x_2, y_2), (x_3, y_3)$ are the coordinates of the corners of a triangle in counterclockwise order then the area of the triangle is \[\textstyle\frac{1}{2}(x_1y_2 - x_1y_3 + x_2y_3 - x_2y_1 + x_3y_1 - x_3y_2).\]

>>>>

We now define the notion of an admissible diagram, which imposes certain density and non-crossing conditions on the configuration of the propagators. The motivation for these conditions comes from the link to SYM $N=4$ theory, where the admissible Wilson loop diagrams are exactly those that correspond to $N^kMHV$ diagrams; see for example \cite[section 2]{HeslopStewart}.

@@

There are a few things to note here.  First, what we call weakly admissible here is called admissible in (\cite{wilsonloop}, see definition 1.11 and section 1.3).  \hlfix{Many of the results below still apply when the hypothesis of admissibility is changed to weak admissibility.}{Propagate "weakly" through document}

@@

As described in the introduction, Wilson loop diagrams come from the study of SYM $N = 4$ in the physics literature. In particular (see \cite[section 2]{HeslopStewart} for details) this gives that each diagram $W$ is associated to a matrix of data $C(W)$, which we now describe.

@@

Matroid theory relates to the study of Wilson loop diagrams as follows. In \cite[section 3]{wilsonloop}, Agarwala and Marin-Amat show that every weakly admissible Wilson loop diagram $W$ with $k$ propagators defines a positroid $M(W)$ of rank $k$ on a base set of size $n$. \hlfix{Combined with Postnikov's work in {\cite{Postnikov}}, this shows that each admissible Wilson loop diagram corresponds to a convex subset of $\Gr(k,n)$}{fix citation. Ask Cameron?}, thus connecting the geometry of Wilson loop diagrams with the geometry underlying the Amplituhedron.

@@

\emph{Polygon dissection} is a common term in combinatorics for a decomposition of a convex polygon into smaller polygons by means of noncrossing diagonals; \hlfix{see for example {\cite{PSpolygon}}}{Karen, can you specify the citation?}.  Each $\tau(W)$ is such an object.

@@

\begin{proof}
Let us first record a few elementary facts about polygon triangulations (that is, about triangulations with all vertices on the outer face).  If such a triangulation has $n$ vertices then it has $n$ edges on the polygon (that is, on the outer face) and $n-3$ edges which are not.  No planar simple graph with the same vertices and the same outer face can have more edges than a triangulation, and every such simple graph with $n-3$ edges off the outer face is a triangulation.\sanote{Karen, can you specify citation in footnote?} \footnote{The last sentence is the statement that maximal outerplanar graphs are exactly polygon triangulations \cite{MPhD}, then the earlier facts follow from Euler's formula for planar graphs.}

@@

In \cite[Theorem 3.39]{wilsonloop}, the authors show that each weakly admissible Wilson loop diagram corresponds to a positroid. That is, each diagram corresponds to a matroid that can be represented by elements of the positive Grassmannian $\Gr(|\cP|, n)$. Any positroid of rank $k$ on $n$ elements defines a subspace of the positive Grassmannian $\Gr(k, n)$, namely the points which represent it \cite{Postnikov}. These subspaces give a CW structure on $\Gr(k,n)$, with each positroid defining a cell.  In other words, there is a map from weakly admissible Wilson loop diagrams $W$ to matroids $M(W)$ to positroid cells in the CW complex on $\Gr(|\cP|, n)$.

@@

Now we are ready to prove the main result of this section, namely that two Wilson loop diagrams define the same matroid (positroid) if and only if they are equivalent (Theorem \ref{same matroid iff equiv}). We also give a formula for the number of Wilson loop diagrams in a given equivalence class (Corollary \ref{number of equiv diagrams}), completing the characterization of the correspondence between Wilson loop diagrams and positroids started in \cite{wilsonloop}.

@@

We know from examples (e.g. \cite[Remark 4.2]{casestudy}) that many positroid cells cannot be obtained as the matroids of admissible Wilson loop diagrams. We do not obtain a formula for the number of cells in the image of $M$ here, but in the next section we give an alternate characterization of ${\rm im}(M)$ in terms of associahedra, with the hope that this linkage with a well-studied combinatorial object will shed light on the issue in the future.

@@

It turns out that this defines a class of combinatorially equivalent polytopes, which can be proved by giving a realization.  \hlfix{There are many interesting realizations of the associahedron {\cite{CSZinequivalent}}}{Karen, can you specify the citation?}. The following realization will be the one that we use.

@@

    Since $\tau(W)$ and $\tau(X)$ do not have the same set of non-triangle faces, there must be some diagonal $d$ of $\tau(W)$ that crosses through a non-triangle face $f$ of $\tau(X)$. 
    It will be most convenient to consider faces in terms of their normal vectors. We make use of the construction discussed on page 8 of \cite{CSZinequivalent}: namely, pick one side of $d$ to be positive and define the $i$th coordinate of the vector $\omega^+_d$ to be
    \[
    \omega^+_d(i) = \begin{cases} \text{dist}(x_i, d) & \text{if $x$ is on the positive side of $d$,}\\ 0 & \text{otherwise.} \end{cases}
    \]
    \hlfix{We have from {\cite{CSZinequivalent}}}{specify citation} that $\omega^+_d$ is a normal vector for any face with $d$ as a diagonal.  However we can also check this fact in a completely elementary way using the shoelace formula for the area of a triangle.  The shoelace formula says that if $(x_1, y_1), (x_2, y_2), (x_3, y_3)$ are the coordinates of the corners of a triangle in counterclockwise order then the area of the triangle is \[\textstyle\frac{1}{2}(x_1y_2 - x_1y_3 + x_2y_3 - x_2y_1 + x_3y_1 - x_3y_2).\]
